---
title: "class08"
format: pdf
author: Whitney Tran (PID:A16781338)
---

Before we get stuck into project work we will have a quick look at applying PCA to some example RNAseq data.


```{r}
url2 <- "https://tinyurl.com/expression-CSV"

rna.data <- read.csv(url2, row.names=1)

head(rna.data)

```

>Q. How many genes are in this dataset?

```{r}
nrow(rna.data)
```

```{r}
##Again we have to take the transpose of our data
pca <- prcomp(t(rna.data), scale=TRUE)

## simple unpolished plot of pc1 and pc2

plot(pca$x[,1], pca$x[,2], xlab="PC1", ylab="PC2")
```
```{r}
summary(pca)
```

```{r}
#we have 5 wt and w ko samples
mycols <- c(rep("blue", 5), rep("red", 5))
mycols
```

```{r}
plot(pca$x[,1], pca$x[,2], xlab="PC1", ylab="PC2", col=mycols)
```


#Analysis of Breast Cancer FNA data

```{r}
wisc.df <- read.csv("WisconsinCancer.csv", row.names=1)

head(wisc.df)
```


```{r}
diagnosis <- as.factor(wisc.df$diagnosis)

diagnosis
```

```{r}
wisc.data <- wisc.df[,-1]
head(wisc.data)
```

>Q1. How many observations are in this dataset?

```{r}
nrow(wisc.df)
```

- 569 observations in this dataset.

>Q2, How many of the observations have a malignant diagnosis?

```{r}
table(wisc.df$diagnosis)
```

- 212 observations have a malignant diagnosis.

>Q3. How many variables/features in the data are suffixed with _mean?

```{r}
colnames(wisc.data)
```

```{r}
length(grep("_mean", colnames(wisc.data), value=T))
```

- 10 variables/features are suffized with "_mean".

#Principal Component Analysis

```{r}
colMeans(wisc.data)
```

```{r}
apply(wisc.data, 2, sd)
```

These are different so we should have scale=TRUE

```{r}
wisc.pr <- prcomp(wisc.data, scale=T)
summary(wisc.pr)
```

>Q.4 From your results, what proportion of the original variance is captured by the first principal components (PC1)?

- 44.27% of the variance is captured by PC1.

>Q5. How many of the principal ocmponents (PCs) are required to describe at least 70% of the original variance in the data?

- 3 principal components are required to describe at least 70%. This is because PC3 gives 72.6% of the variance whereas PC2 only gives 63.2%.

>Q6. How many principal components (PCs) are required to describe at least 90% of the original variance in the data?

- 7 principal components are required to describe at last 90% of the original variance. (PC7 covers 91.0% of the original variance).

#Plotting the PCA results

```{r}
biplot(wisc.pr)
```


>Q7. What stands out to you about this plot? Is it easy or difficult to understand? Why?

- I find it difficult to read and understand. The numbers, words, and colors are stacked on top of each other making it look like a lot of noise rather than a defined plot.

```{r}
attributes(wisc.pr)
```
```{r}
plot(wisc.pr$x[,1], wisc.pr$x[,2], col=diagnosis)
```

>Q8. Generate a similar plot for principal components 1 and 3. What do you notice about these plots?

```{r}
plot(wisc.pr$x[,1], wisc.pr$x[,3], col=diagnosis)
```

- PC1 provides a separation of malignant from benign samples. PC3 shows less variance than PC2.

```{r}
library(ggplot2)

pc<- as.data.frame(wisc.pr$x)

ggplot(pc) + aes(PC1, PC2, col=diagnosis) + geom_point()
```

```{r}
#Calculate the variance of each component

pr.var <- wisc.pr$sdev^2

head(pr.var)
```

#Communicating PCA Results

>Q9. For the first principal component, what is the component of the loading vector (i.e wisc.pr$rotation[,1]) for the feature concavepoints_mean?

```{r}
wisc.pr$rotation["concave.points_mean",1]
```

- -0.26 is the component of the loading vector for concave.points_mean.

>Q10. What is the minimum number of principal components required to expain 80% of the variance of the data?

```{r}
tbl <- summary(wisc.pr)
which(tbl$importance[3,] > 0.8)[1]
```

- PC5 is the minimum number of principal components to explain 80% of the variance.

#Hierarchical Clustering

>Q11. Using the plot() and abline() functions, what is the height at which the clustering model has 4 clusters?

```{r}
d <- dist(scale(wisc.data))
wisc.hclust <- hclust(d)
plot(wisc.hclust)
```
```{r}
plot(wisc.hclust)
abline(h=18, col="red")
```
```{r}
grps <- cutree(wisc.hclust, h=18)
table(grps)
```

- It has 4 clusters at a height of 18.

section 5, skip 4

#5. Combining MEthods

We will perform clustering on our PCA results rather than the original data.

In other words, we will cluster using `wisc.pr$x` - our new better variables or PCs. We can choose as many or as few PCs to use as we like.

```{r}
d.pc <- dist(wisc.pr$x[,1:3])
wisc.pr.hclust <- hclust(d.pc, method = "ward.D2")
plot(wisc.pr.hclust)
abline(h=80, col="red")
```

```{r}
grps <- cutree(wisc.pr.hclust, h=80)
table(grps)
```

We can use `table()` function to make a cross-table as well as a count table.

```{r}
table(diagnosis)
```

```{r}
table(grps, diagnosis)
```
The results show that cluster 1 mainly captures malignant cancer whereas cluster 2 caputres relatively health samples.

#Prediction

```{r}
#url <- "new_samples.csv"

url<- "https://tinyurl.com/new-samples-CSV"
new <- read.csv(url)
npc <- predict(wisc.pr, newdata=new)
npc
```

```{r}
plot(wisc.pr$x[,1:2], col=grps)
points(npc[,1], npc[,2], col="blue", pch=16, cex=3)
text(npc[,1], c(1,2), col="white")
```

>Q18. Which of these new patients should we prioritize for follow up based on your results.

- Patient 2 should be prioritized because they are among the cluster that captures more malignant cancer samples.
